Early environmental conditions do not impact behavioural flexibility in an invasive and non-invasive lizard species

### 2024-03-26 ms.qmd
Updates

Pablo Recio-Santiago^1^,$\ddagger$ , Dalton C. Leibold^1^, Ondi Crino ^1,2^, Kris Wild^1,3^, Chris Friesen^4^, Basile Mauclaire^1,5^, Amelia Peardon ^1^, Daniel W.A. Noble^1^

@@

## Abstract
Behavioural flexibility is crucial for species adapting to new environments, specially during invasion events. However, the early environment experienced can profoundly impact animals' ability to behave flexibly. Aspects like parents' transmitted hormones or temperature are predicted to interact to shape behavioural flexibility. But, given the advantages associated to behaving flexibly during the invasion process, behavioural flexibility in invasive species is expected to be less susceptible to the impacts of early environment. Here, we investigated the interactive effects of prenatal corticosterone (CORT) and temperature on behavioural flexibility in two lizard species, _Lampropholis delicata_ and _L. guichenoti_, with distinct invasion success. We manipulated prenatal CORT levels and incubation temperature in a 2x2 factorial design, and then assessed behavioural flexibility through a reversal learning task. Contrary to our expectations, we found no significant differences in the reversal learning rate between treatments or species. Our results suggest that prenatal temperature and CORT may not influence reversal learning in these species and that other factors in _L. delicata_ biology may be more influential in invasion success than behavioural flexibility alone.

@@

New environments faced by invasive populations introduce organisms to a plethora of unfamiliar, and potentially stressful, situatons. Sensory systems detect and transmit environmental signals to instigate adaptive physiological and behavioural adjustments – responses  that are mediated by glucocorticoids (GCs) [@sapolsky_how_2000]. Glucocorticoids can be transmitted to offspring and influence their phenotype and development through transgenerational effects [@Crino_2023]. In addition to novel enviroments experienced by parents, offspring also experience potentially stressful environmental conditions that can interact with or amplify parental effects. Temperature, in particular, is likely to deviate in invasive populations compared to source populations having a direct effect on offspring phenotype with long-lasting consequences [@noble_developmental_2018]. As such, GCs and temperature are predicted to interact to influence the offspring phenotype with particuarly potent effects on brain development [@jonson1976effect; @coomber_independent_1997; @zhu_prenatal_2004; @amiel_effects_2017] and cognition [@clark_colour_2014; @lui2017chronic]. While the importance of behavioural flexibility is recognised as being important for invasion success, we know little about how novel environmental stressors, such as GCs and temperature, impact behavioural flexibility. Such knowledge is crucial for understanding the potential for invasive populations to adapt to new environments and the potential for invasion success.

Here, we studied the effects of prenatal temperature and the glucocorticoid, corticosterone (CORT) (the main GC in reptiles) [@Crino_2023], on behavioural flexibility in two species with distinct histories of invasion success, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). _L. delicata_ has been succesful in colonising different areas around the globe [@baker1979rainbow; @chapple2013divergent; @chapple2015deliinvLHI], but the same is not true of _L. guichenoti_ despite the two species having similar life-histories and overlapping distributions [@chapple_know_2011; @chapple_biology_2014]. Personality has been shown to be different between these two species [@chapple_know_2011; @chapple2012can; @bezzina2014does], which could be related to general cognitive differences that are important for invasion success. Additionally, early environments are known to affect suites of morphological and behavioural traits differently in each species [@carrasco2023maternal; @kar2023heritability], but the interactive role of maternal CORT and early temperatures have not yet been explored. Differences in the relative impact that early environments have on behavioural flexibility may help explain differing invasion success between the two species. We predicted that _L. delicata_ would, on average, perform more proficiently in the task compared to _L. guichenoti_. We also anticipated prenatal CORT and temperature to affect the behavioural flexibility of both species, and that _L. guichenoti_ would be more impacted by the prenatal environment than _L. delicata_.

@@

```{r, fig-Methods}
#| label: fig-Methods
#| fig.cap: "Experimental design of early environment manipulation and learning tasks. On top, the early environment manipulation for both species. Center, the habituation phase with the respective three different stages. On the bottom, the associative and reversal tasks; white lids show the ramps where the food reward was not accessible."

@@

```{r, fig-results}
#| label: fig-results
#| fig.cap: "Results for L. delicata (A,B) and L. guichenoti (C, D). A, C) Predicted probability of choosing the correct feeder first over trials. The lines represent the mean predicted probability of choosing the correct feeder first, and the shaded areas represent the standard deviation of the mean; both obtained by using the slope and intercept estimates from the posterior distributions. The different colours represent the different treatments. B, D) Distribution of the estimates of the slopes per each treatment. The x-axis represents the estimates of the slopes, and the y-axis represents the density of the estimates. The different colours represent the different treatments. Points and bars represent the mean and standard deviation of the mean of the estimates, respectively."
source(here("R", "func.R"))
# Create the dfs for the probability plot A and C) (df_prob)
# L. delicata
df_deliprob <- df_plotAC("deli")
# L. guichenoti
df_guichprob <- df_plotAC("guich")
#
# Create the df for the violin plots in plot B) (df_violin)
# L. delicata
df_deliV <- df_plotBD1("deli")
# L. guichenoti
df_guichV <- df_plotBD1("guich")
#
# Create the df for the points and bars in plot B) (df_points)
# L. delicata
df_deliP <- df_plotBD2("deli")
# L. guichenoti
df_guichP <- df_plotBD2("guich")
#
#
# Make plots deli and guich
source(here("R", "func.R"))
fig_deli <- plotting(sp = "deli", df_prob = df_deliprob, df_violin = df_deliV, df_points = df_deliP)
fig_guich <- plotting(sp = "guich", df_prob = df_guichprob, df_violin = df_guichV, df_points = df_guichP)
# Combine plots of both species into the final figure
fig_results <- plot_grid(fig_deli, fig_guich, ncol = 1)
ggsave("./output/figures/fig_results.png", plot=fig_results, width = 25, height = 18, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/fig_results.png")
```

@@

The study embarked on unravelling the interactive effects of prenatal temperature and GCs on behavioural flexibility in two closely related species, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). Despite sharing life history traits, only _L. delicata_ has proven successful in colonising various global regions, and we hypothesised this may be related to higher behavioural flexibility. Furthermore, we expected behavioural flexibility in this species to be less susceptible to the effect of early environment since this trait can be adaptive in several stages of invasion involving more than one generation. Nevertheless, we did not find significant differences between treatments or species when we compared their performance in a reversal task. We found that, on average, all the lizards from the four treatments and the two species learned the reversal task - indicated by estimated slopes that are significantly different from zero - and that they learn, on average, at the same rate. Furthermore, these results were consistent when only those individuals that performed better in the previous associative task were included (see Supplementary material).  
Previous studies have shown that _L. delicata_ more readily explores novel environments than _L. guichenoti_, which can be advantageous during colonisation [@chapple_know_2011]. However, Bezzina et al. [@bezzina2014does] found no differences in learning between the two skinks in an associative learning task. Our results align with the latter study and suggest that behavioural flexibility does not differ between these species. In addition, we did not find any effect of the treatments on either species. Both results contrast our original predictions and may indicate that other aspects of their biology are more relevant for _L. delicata_'s success as invaders than behavioural flexibility. For instance, the mentioned differences in exploratory behaviour can be the key factor explaining why _L. delicata_ and not _L. guichenoti_ is better at colonising new areas, since exploring more can also enhance the ability to effectively locate novel resources in _L. delicata_ [@chapple_know_2011].

Our early environment manipulation did not affect the lizards' performance in the task. We predicted that prenatal CORT treatment would decrease reversal learning abilities and that high temperatures would minimise these effects. However, our results show no significant differences between treatments in both species. Our outcome contrasts with previous studies that show impairments of high GCs levels in different types of learning tests [@lemaire_prenatal_2000; @zhu_prenatal_2004;  @farrell_developmental_2016], but they are consistent with other experiments where GCs increased or did not impact learning abilities [@szuran_water_1994; @bebus_associative_2016]. More striking is the absence of an effect of incubation temperature since most of the studies in reptiles reported better performance when incubated at higher temperatures in different associative learning and motor tasks [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]. Furthermore, those studies that reported cold-incubated lizards learn faster [@dayananda_incubation_2017; @abayarathna_effects_2020] used incubation temperatures far beyond the natural range in the hot treatment. In some systems, exposure to CORT during development has hormetic effects on  learning and neural structures [@du_dynamic_2009; @mcewen_brain_2012]. For example, in Sprague–Dawley rats, exposure to high doses of CORT, but not low doses, reduced density of immature neurons and cell proliferation and in the dentate gyrus [@DOSEneurobrummelte2010chronic]. We know that prenatal CORT treatment affects other traits in both species of lizards used in this experiment (unpublished data). However, it is possible that prenatal CORT treatment affects tissues differently. Conversely, it is likely that the effects of the early environment may vary depending on the cognitive task, as evidenced by studies showing variations in the impact of prenatal GCs and temperature on different brain regions [@coomber_independent_1997; @sakata_neural_2000; @lemaire_prenatal_2000]. These hypotheses are not mutually exclusive and could be tested in future experiments using multiple doses of prenatal CORT treatment, assessing learning using multiple cognitive tasks, and by measuring the neurophysiological effects on different brain nuclei.

>>>>

Pablo Recio-Santiago^1^,$\ddagger$ , Dalton C. Leibold^1^, Ondi Crino ^1,2^, Kristoffer H. Wild^1,3^, Chris Friesen^4^, Basile Mauclaire^1,5^, Amelia Peardon ^1^, Daniel W.A. Noble^1^

@@

## Abstract
Behavioural flexibility is crucial for species adapting to new environments, specially during invasion events. However, the early environment experienced can profoundly impact animals' ability to behave flexibly. Aspects like developmental environments and parental investment are predicted to interact to shape behavioural plasticity. However, given the advantages associated to behaving flexibly during the invasion process, behavioural flexibility in invasive species is expected to be less susceptible to the impacts of early environment. Here, we investigated the interactive effects of prenatal corticosterone (CORT) and temperature on behavioural plasticity in two lizard species, _Lampropholis delicata_ and _L. guichenoti_, with distinct histories of invasion. We manipulated prenatal CORT levels and incubation temperature in a 2x2 factorial design, and then assessed behavioural flexibility through a reversal learning task. Contrary to our expectations, we found no significant differences in the reversal learning rate between treatments or species. We argue that prenatal temperature and CORT may not influence reversal learning in these species, which suggests that other aspects of _L. delicata_ biology may be more influential in invasion success than behavioural flexibility alone.

@@

Novel environments faced by invasive populations introduce organisms to a plethora of unfamiliar and potentially stressful situatons. Sensory systems detect and transmit environmental signals to instigate adaptive physiological and behavioural adjustments – responses  that are mediated by glucocorticoids (GCs) [@sapolsky_how_2000]. Glucocorticoids can be transmitted to offspring and influence their phenotype and development through transgenerational effects [@Crino_2023]. In addition to novel environments experienced by parents, offspring also experience potentially stressful environmental conditions that can interact with or amplify parental effects. Temperature, in particular, is likely to deviate in invasive populations compared to source populations, which have a direct effect on offspring phenotypes and long-lasting consequences [@noble_developmental_2018]. As such, GCs and temperature are predicted to interact to influence the offspring phenotype with particularly potent effects on brain development [@jonson1976effect; @coomber_independent_1997; @zhu_prenatal_2004; @amiel_effects_2017] and cognition [@clark_colour_2014; @lui2017chronic]. While behavioural flexibility is recognised as an important behavioural phenotype for invasion success [@wright2010behavioral; @chapple2012can; @szabo2020can], we know little about how novel environmental stressors, such as GCs or temperature, impact behavioural flexibility. Such knowledge is crucial for understanding the potential for invasive populations to adapt to new environments and the potential for invasion success.

Here, we studied the effects of prenatal temperature and the glucocorticoid, corticosterone (CORT) (the main GC in reptiles) [@Crino_2023], on behavioural flexibility in two species with distinct histories of invasion success, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). _Lampropholis delicata_ has been successful in colonising different areas around the globe [@baker1979rainbow; @chapple2013divergent; @chapple2015deliinvLHI], but the same is not true of _L. guichenoti_ despite the two species having similar life-histories and overlapping distributions [@chapple_know_2011; @chapple_biology_2014]. Personality has been shown to be different between these two species [@chapple_know_2011; @chapple2012can; @bezzina2014does], which could be related to general cognitive differences that are important for invasion success. Additionally, early environments are known to affect suites of morphological and behavioural traits differently in each species [@carrasco2023maternal; @kar2023heritability], but the interactive role of maternal CORT and early temperatures has not yet been explored. Differences in the relative impact of early environments on behavioural flexibility may help explain differing invasion success between the two species. We predict that _L. delicata_, on average, will perform more proficiently in the task compared to _L. guichenoti_. We also anticipate prenatal CORT and temperature to affect the behavioural flexibility of both species, and that _L. guichenoti_ will be more impacted by the prenatal environment than _L. delicata_.

@@

```{r, fig-Methods}
#| label: fig-Methods
#| fig.cap: "Experimental design of early environment manipulation and learning tasks. Panel A represents the early environment manipulation for both species. Panel B shows the habituation phase with the respective three different stages. And panel C represents the associative and reversal tasks; white lids show the ramps where the food reward was not accessible."

@@

```{r, fig-results}
#| label: fig-results
#| fig.cap: "Results for L. delicata (A,B) and L. guichenoti (C, D). Panels A and C show the predicted probability of choosing the correct feeder first over trials. The lines represent the mean predicted probability of choosing the correct feeder first on each trial, and the shaded areas indicate the standard deviation of the mean; both were obtained by using the slope and intercept estimates from the posterior distributions. The different colours indicate the different treatments. Panels B and D show the distribution of the estimates of slopes per each treatment. The x-axis represents the slope estimate, and in the y-axis are the density of the estimates. The different colours indicate the different treatments. Points and bars represent the mean and standard deviation of the mean of the estimates, respectively."
source(here("R", "func.R"))
# Create the dfs for the probability plot A and C) (df_prob)
# L. delicata
df_deliprob <- df_plotAC("deli")
# L. guichenoti
df_guichprob <- df_plotAC("guich")
#
# Create the df for the violin plots in plot B) (df_violin)
# L. delicata
df_deliV <- df_plotBD1("deli")
# L. guichenoti
df_guichV <- df_plotBD1("guich")
#
# Create the df for the points and bars in plot B) (df_points)
# L. delicata
df_deliP <- df_plotBD2("deli")
# L. guichenoti
df_guichP <- df_plotBD2("guich")
#
#
# Make plots deli and guich
source(here("R", "func.R"))
fig_deli <- plotting(sp = "deli", df_prob = df_deliprob, df_violin = df_deliV, df_points = df_deliP)
fig_guich <- plotting(sp = "guich", df_prob = df_guichprob, df_violin = df_guichV, df_points = df_guichP)
# Combine plots of both species into the final figure
fig_results <- plot_grid(fig_deli, fig_guich, ncol = 1)
ggsave("./output/figures/fig_results.png", plot=fig_results, width = 25, height = 18, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/fig_results.png")
```

@@

The study aimed to determine the interactive effects of prenatal temperature and GCs on behavioural flexibility in two closely related species, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). Despite sharing life history traits, only _L. delicata_ has proven successful in colonising various global regions [@baker1979rainbow; @chapple2013divergent; @chapple2015deliinvLHI], and we predicted this may relate to higher behavioural flexibility. Furthermore, we expected behavioural flexibility in this species to be less susceptible to the effect of early environment since this trait can be adaptive in several stages of invasion involving more than one generation. Nevertheless, we did not find significant differences between treatments or species when we compared their performance in a reversal task. We found that, on average, all the lizards from the four treatments and the two species learned the reversal task - indicated by estimated slopes that are significantly different from zero - and that they learn, on average, at the same rate. Furthermore, these results were consistent when only those individuals who performed better in the previous associative task were included (see Supplementary material).

Previous studies have shown that _L. delicata_ more readily explores novel environments than _L. guichenoti_, which can be advantageous during colonisation [@chapple_know_2011]. However, Bezzina _et al._ [@bezzina2014does] found no differences in learning between the two skinks in an associative learning task. Our results align with the latter study and suggest that behavioural flexibility does not differ between these species. In addition, we did not find any effect of the treatments on either species. Both results contrast our original predictions and may indicate that other aspects of their biology are more relevant for _L. delicata_'s success as invaders than behavioural flexibility. For instance, the mentioned differences in exploratory behaviour can be the key factor explaining why _L. delicata_ and not _L. guichenoti_ is better at colonising new areas, since exploring more can also enhance the ability to effectively locate novel resources in _L. delicata_ [@chapple_know_2011].

Our early environment manipulation did not affect the lizards' performance in the task. We predicted that prenatal CORT treatment would decrease reversal learning abilities and that high temperatures would minimise these effects. However, our results show no significant differences between treatments in both species. In contrast to previous studies that show high GCs levels impaired performance in different types of learning tests [@lemaire_prenatal_2000; @zhu_prenatal_2004;  @farrell_developmental_2016], we did not find any effect of CORT on the speed at which lizards learn. Our results are, however, consistent with other experiments where GCs did not impact learning abilities [@szuran_water_1994; @bebus_associative_2016]. More striking is the absence of an effect of incubation temperature since most of the studies in reptiles reported better performance when incubated at higher temperatures in different associative learning and motor tasks [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]. Furthermore, those studies that reported cold-incubated lizards learn faster [@dayananda_incubation_2017; @abayarathna_effects_2020] used incubation temperatures far beyond the natural range in their hot treatments. In some systems, exposure to CORT during development has hormetic effects on  learning and neural structures [@du_dynamic_2009; @mcewen_brain_2012]. For example, in Sprague–Dawley rats, exposure to high doses of CORT, but not low doses, reduced density of immature neurons and cell proliferation and in the dentate gyrus [@DOSEneurobrummelte2010chronic]. We know that prenatal CORT treatment affects other traits in both species of lizards used in this experiment (unpublished data). However, it is possible that prenatal CORT treatment affects tissues differently. Conversely, the effects of the early environment may likely vary depending on the cognitive task, as evidenced by studies showing variations in the impact of prenatal GCs and temperature on different brain regions [@coomber_independent_1997; @sakata_neural_2000; @lemaire_prenatal_2000]. These hypotheses are not mutually exclusive and could be tested in future experiments using multiple doses of prenatal CORT treatment, assessing learning using multiple cognitive tasks, and measuring the neurophysiological effects on different brain nuclei.
